A novel method for detecting coalescing binaries in near realtime with Advanced LIGO and beyond \editorial{Submit to PRD or CQG?}

### 2011-02-15 introduction.tex
lloid_paper intro: add more science and speculate on negative latency
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@630 ccdd05d5-56c1-461a-bf98-85f6876523b4

Low-mass inspiraling binaries are the most promising sources of gravitational
radiation for Advanced \textsc{ligo}. Observation demonstrates that they must
exist in the local Universe in some abundance~\cite{Abadie:2010p10836}.
[science] In the
final stages of inspiral, the neutron star is tidally disrupted, providing the
matter required to produce bright electromagnetic
counterparts~\cite{shibata:2007}. Conveniently, the gravitational waveforms are
understood and implemented to the level required by Advanced \textsc{ligo} to
detect~\cite{BuonannoIyerOchsnerYiSathya2009}, allowing the sensitive technique
of matched filtering.

The parameter space of compact binary coalescence signals is
large~\cite{Owen:1995tm, Owen:1998dk} leading to large filter banks and
significant computational cost. Following the series of detector upgrades that
are now under way, \textsc{ligo} and Virgo will attain their Advanced
configurations, gaining a tenfold improvement in amplitude sensitivity over
Initial configurations~\cite{advLIGOrefdesign,advVirgorefdesign} and a
corresponding thousandfold increase in observable volume in the local Universe.
The computational requirements of conventional detection strategies will be
greatly increased by this upgrade for two reasons. First, advanced detectors
will have much improved low-frequency sensitivity, so chirp signals from
coalescing binaries will remain in the detectors' band for a significantly
longer time. This will demand that matched filter based searches use longer
templates, requiring more memory and more cycles to process. Second, background
noise for advanced detectors will be more spectrally uniform, so it will be
possible to resolve the intrinsic parameters of a source that determine the
time-frequency structure of the gravitational wave signal with much greater
accuracy. This will require that matched filter bank based searches employ
more matched filters, again demanding more memory and more cycles.

@@

In \oldstylenums{2010}, \textsc{ligo} and Virgo completed S6/VSR3,
a short period of
joint data taking during which several all-sky detection pipelines operated in
a low-latency configuration \citeneeded. A few candidates of moderate
significance were promptly sent for electromagnetic followup \citeneeded{} to
several telescopes including Swift, \textsc{rotse}, \textsc{tarot}, and Zadko.
They achieved latencies of BLAH, but required human vetting of each candidate
before alerting the telescope partners. As confidence in the infrastructure
increases, humans can be removed from the loop, setting a demonstrated previous
best latency at BLAH.

>>>>

Low-mass inspiraling binaries are the most promising sources of gravitational
radiation for Advanced \textsc{ligo}. Observation demonstrates that they must
exist in the local Universe in some abundance~\cite{Abadie:2010p10836}.
A detection would strongly constrain mass and
spin~\cite{finn1993,Poisson:1995ef}, and with sufficient signal strength, the
neutron star equation of state~\cite{flanagan:021502,Read:2009}, general
relativity in the strong-field regime~\cite{Will:2005va}, the graviton
mass~\cite{PhysRevD.80.044002,Keppel:2010qu}, and the Lorentz Invariance
principle~\cite{Ellis2006402}. In the
final stages of inspiral, the neutron star is tidally disrupted, providing the
matter required to produce bright electromagnetic
counterparts~\cite{shibata:2007}. Conveniently, the gravitational waveforms are
understood and implemented to the level required by Advanced \textsc{ligo} to
detect~\cite{BuonannoIyerOchsnerYiSathya2009}, allowing the sensitive technique
of matched filtering.

The parameter space of compact binary coalescence signals is
large~\cite{Owen:1995tm, Owen:1998dk} leading to large filter banks and
significant computational cost. Following the series of detector upgrades that
are now under way, \textsc{ligo} and Virgo will attain their Advanced
configurations, gaining a tenfold improvement in amplitude sensitivity over
Initial configurations~\cite{advLIGOrefdesign,advVirgorefdesign} and a
corresponding thousandfold increase in observable volume in the local Universe.
The computational requirements of conventional detection strategies will be
greatly increased by this upgrade for two reasons. First, advanced detectors
will have much improved low-frequency sensitivity, so chirp signals from
coalescing binaries will remain in the detectors' bands for a significantly
longer time. This will demand that matched filter based searches use longer
templates, requiring more memory and more cycles to process. Second, background
noise for advanced detectors will be more spectrally uniform, so it will be
possible to resolve the intrinsic parameters of a source that determine the
time-frequency structure of the gravitational wave signal with much greater
accuracy. This will require that matched filter bank based searches employ
more matched filters, again demanding more memory and more cycles.

@@

\editorial{More speculative and no reference. Thoughts? Move to discussion? Worth adding any plots?}
More tantalizingly, these long waveforms enable us to accumulate a significant
fraction of the signal-to-noise ratio before coalescence. With enough
signal-to-noise ratio (and a lot of work!) it should be possible to reconfigure
the detectors on-the-fly to more sensitively capture the later merger part of
the signal, where neutron star equation of state information can be extracted.
With similar caveats, it is certainly possible to send alerts to telescopes
several seconds before prompt emission. There are no theoretical hurdles to
negative latency, but careful attention must be paid to the algorithms and
implementation of search pipelines to take advantage of a fortuitously loud
signal.

\editorial{If we publish before the LUMIN papers, we have to redo this section.
We can't make claims about those private communications with the telescopes.
Else, cite the pre-S6 LUMIN paper \cite{kanner2008}.}
In \oldstylenums{2010}, \textsc{ligo} and Virgo completed S6/VSR3,
a short period of
joint data taking during which several all-sky detection pipelines operated in
a low-latency configuration \citeneeded. A few candidates of moderate
significance were promptly sent for electromagnetic followup \citeneeded{} to
several telescopes including Swift, \textsc{rotse}, \textsc{tarot}, and Zadko.
They achieved latencies of BLAH, but required human vetting of each candidate
before alerting the telescope partners. As confidence in the infrastructure
increases, humans can be removed from the loop, demonstrating a best latency
of BLAH.
